Comparison of Fast, Hybird Imaging Architectures for Multi-scale, Hierarchical Aperture Arrays

### 2024-06-26 ms.tex
Update on Overleaf.

\begin{abstract}
Two major areas of modern radio astronomy, namely, explosive astrophysical transient phenomena and observational cosmology, are driving aperture arrays towards large-$N$, low-cost element design consisting of multiple spatial scales. Multi-scale, hierarchical aperture arrays require a combination of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different spatial scales to obtain optimal performance in imaging the entire field of view. We use the metric of computationalﬁ cost density defined over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time to determine the most efficient hybrid architecture for aperture arrays spanning multiple spatial scales like the element size, stations (a collection of elements), and the array (a collection of stations) over a range of number of elements per station and number of stations. We use nominal parameters of specific upcoming and planned arrays-- Square Kilometre Array (SKA) at low frequencies (\texttt{SKA-low}), \texttt{SKA-low-core}, a long baseline extension to SKA-low (\texttt{LAMBDA-I}), compact all-sky phased array (\texttt{CASPA}), and a lunar array (\texttt{FarView-core})-- to determine the most optimal architectural hierarchy for each from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large-$N$, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval.
\end{abstract}

@@

Here we describe the first stage, namely, intra-station architectures that form the basis for both approaches. This first stage essentially synthesises station-scale apertures with simultaneous, multiple virtual pointings. These station-scale synthesised apertures from the first stage will act as building blocks for the second stage of inter-station processing. We will use $a_m$ to denote element, $a$, in station, $m$.

@@

Figure~\ref{fig:1D-incoherent-compcost-BF-LAMBDA-I} shows the breakdown of various components of the computational cost per station per voxel for voltage beamforming as a function of the different station parameters. In each panel, all parameters except the one on the corresponding $x$-axis are kept fixed at the characteristic values of the \texttt{LAMBDA-I} station as highlighted in cyan. The total cost is dominated by the beamforming in Equation~(\ref{eqn:intra-station-pol-hol-img-expl}) denoted by black dashed lines. Beamforming cost per pixel scales linearly with the number of elements per station. The costs of squaring and temporal averaging in Equation~(\ref{eqn:intra-station-opt-pol-img-outprod}) are subdominant. Because they are performed on a per-pixel basis and not dependent on $N_\textrm{a/s}$, they remain constant across $N_\textrm{a/s}$. And because the number of independent pixels covering the field of view scales as $n_\textrm{bs} \simeq (D_\textrm{s}/D_\textrm{a})^2$, the cost per pixel remains constant across $D_\textrm{s}$ and $D_\textrm{a}$. All these operations have to be performed at every time step, $\delta t$, and hence remain constant regardless of $t_\textrm{acc}$. Although spatial averaging across multiple stations (orange dot-dashed line) is not a station-level operation, it is shown only to emphasise that constitutes a negligible cost relative to the other components. Note that it needs to be only performed on a slower cadence of $t_\textrm{acc}$ on a per-pixel basis, and thus scales inversely with $t_\textrm{acc}$.

@@

Figure~\ref{fig:multidim-incoherent-compcost-BF-LAMBDA-I} shows a covariant view of the total computational cost density per station per voxel for the beamforming architecture taken two station parameters at a time while keeping the rest fixed at the nominal values of a \texttt{LAMBDA-I} station denoted by grey dashed lines. The variation in the cost per voxel is dominated by the beamforming cost and displays the same trends observed in Figure~\ref{fig:1D-incoherent-compcost-BF-LAMBDA-I}, namely, linear in $N_\textrm{a/s}$ while remaining constant with $D_\textrm{a}$ and $D_\textrm{s}$.

@@

A simultaneous beamforming filling the entire field of view can also be achieved by a two-dimensional spatial Fast Fourier Transform (FFT) of the measurements on the aperture plane. This can be enabled even for an irregular layout of station elements by gridding the calibrated electric fields measured by the station elements. The Optimal Map Making formalism \citep[OMM;][]{Tegmark1997a} formalism adopted by \citet{Morales2011} is the basis of the architecture called E-field Parallel Imaging Correlator \cite[EPIC;][]{Thyagarajan+2017},
\begin{align}
    \widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k) &= \sum_j \delta^2 \boldsymbol{r}_j \, e^{i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_j} \left(\sum_{a_m} \sum_p \widetilde{W}_{a_m}^{\alpha p*}(\boldsymbol{r}_j-\boldsymbol{r}_{a_m}) \, \widetilde{E}_{a_m}^p \right) \, . \label{eqn:intra-station-pol-hol-img-epic}
\end{align}
The expression within the parenthesis represents the gridding operation, and the summation outside denotes the Fourier transform. The polarised intensities are estimated according to Equation~(\ref{eqn:intra-station-opt-pol-img-outprod}).

@@

Application of the FFT will have the effect of simultaneously beamforming over the entire field of view, $\Omega_\textrm{a}$. The convolution with $\widetilde{W}_{a_m}^{p\alpha*}(\boldsymbol{r})$ has two purposes. Firstly, it can be chosen to optimise specific properties in the synthesised image. For example, choosing $\widetilde{W}_{a_m}^{p\alpha}(\boldsymbol{r})=W_{a_m}^{p\alpha*}(\boldsymbol{r})$ will optimise the signal-to-noise ratio in the image. Secondly, it also acts as gridding convolution transforming data from discrete and arbitrary locations onto a regular grid, thereby facilitating the application of a two-dimensional spatial FFT.

@@

Figure~\ref{fig:1D-incoherent-compcost-EPIC-LAMBDA-I} shows a breakdown of the costs per station per voxel of various components constituting the EPIC architecture, with nominal values of a \texttt{LAMBDA-I} station marked in cyan. The dominant cost is from the two-dimensional FFT (black dashed lines) in Equation~\ref{eqn:intra-station-pol-hol-img-epic}.
% or equivalently, Equation~\ref{eqn:intra-station-pol-hol-img-epic-LA}.
Because the spatial FFT cost depends only on the grid dimensions, it is independent of $N_\textrm{a/s}$ as the grid can hold any $N_\textrm{a/s}$ elements as physically possible. This spatial FFT scales roughly as $(\gamma_\textrm{s} \, D_\textrm{s}/D_\textrm{a})^2\log_2(\gamma_\textrm{s}\, D_\textrm{s}/D_\textrm{a})^2$, where, $\gamma_\textrm{s}$ is a constant padding factor. Notably, because the spatial FFT produces an image over the entire field of view at every independent pixel in one instant, it is not a pixelwise operation, and therefore, its cost per pixel is $\approx \log_2(D_\textrm{s}/D_\textrm{a})$, showing a weak dependence on $D_\textrm{s}$ and $D_\textrm{a}$. The pixelwise squaring (pink dot-dashed lines) and temporal averaging (blue dot-dashed lines) costs in Equation~\ref{eqn:intra-station-opt-pol-img-outprod}
% {eqn:intra-station-opt-polarimetric-direct-img-LA}
are subdominant. Being per pixel operations, the cost per voxel remains constant with $N_\textrm{a/s}$, $D_\textrm{s}$ and $D_\textrm{a}$ similar to the voltage beamforming architecture. The gridding (black dotted lines) with weights in Equation~(\ref{eqn:intra-station-pol-hol-img-epic})
% and (\ref{eqn:intra-station-pol-hol-img-epic-LA})
is a per-element operation and depends only on $N_\textrm{a/s}$, it scales linearly with $N_\textrm{a/s}$ and inverse squared with $D_\textrm{s}$ per voxel. Overall, the gridding cost is significantly subdominant. Again, the spatial averaging of pixels across multiple stations (orange dot-dashed line) is not a station-level operation, but is only shown to emphasise that it operates on a much slower cadence of $t_\textrm{acc}$, scaling inversely with $t_\textrm{acc}$, and adds only negligible cost to the overall cost budget.

@@

\begin{figure*}
\centering
\subfloat[][Covariant cost density of EPIC architecture \label{fig:multidim-incoherent-compcost-EPIC-LAMBDA-I}]{\includegraphics[width=0.48\textwidth]
% {figures/multidim_epic_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_05a.pdf}  
}
\subfloat[][Covariant cost of EPIC architecture relative to BF architecture \label{fig:multidim-incoherent-relative-compcost-EPIC-LAMBDA-I}]{\includegraphics[width=0.48\textwidth]
% {figures/multidim_epic_relative_cost_analysis_LAMBDA-I.pdf}
{fig_05b.pdf}
} 
\caption{(Left): Two-dimensional slices of the computational cost per station per voxel for coherent-incoherent imaging using station-level EPIC. Because the spatial FFT dominates the overall cost budget, the computational cost density is relatively insensitive to $N_\textrm{a/s}$, $D_\textrm{s}$, and $D_\textrm{a}$. (Right): Ratio of computational cost of EPIC to voltage beamforming. For $N_\textrm{a/s}\gtrsim 100$, EPIC has a significant advantage over voltage beamforming Cyan contours indicate nominal values for a \texttt{LAMBDA-I} station. \label{fig:incoherent-compcost-EPIC-LAMBDA-I}}
\end{figure*}

@@

Figure~\ref{fig:multidim-incoherent-compcost-XBF-LAMBDA-I} shows the covariant view of the total computational cost density per station per voxel for the correlator beamformer architecture taken two station parameters at a time while keeping the rest fixed at the nominal values of a \texttt{LAMBDA-I} station denoted by grey dashed lines. The variation in the cost per voxel for $t_\textrm{acc}=1$~ms is dominated by the spatial DFT cost which is relatively constant in $D_\textrm{a}$ and $D_\textrm{a}$ and scales as $N_\textrm{a/s}^2$, displaying the same trends observed in Figure~\ref{fig:1D-incoherent-compcost-XBF-LAMBDA-I}. Figure~\ref{fig:multidim-incoherent-relative-compcost-XBF-LAMBDA-I} shows the ratio of the computational cost of the XBF architecture relative to the voltage beamforming architecture in Figure~\ref{fig:multidim-incoherent-compcost-BF-LAMBDA-I}. It is clearly noted that the cost of the XBF architecture is less expensive only for $N_\textrm{a/s}\lesssim 100$ and gets worse with increasing $N_\textrm{a/s}$.

@@

In this approach, $\widetilde{V}_{a_m b_m}^{pq}$, the calibrated and inverse noise covariance weighted station visibilities in station $m$ in Equation~(\ref{eqn:intra-station-pol-visibilities}) are gridded with weights, $\widetilde{B}_{a_m b_m}^{\alpha\beta pq*}(\Delta\boldsymbol{r}_j-\Delta\boldsymbol{r}_{a_m b_m})$, onto a grid and Fourier transformed using an FFT rather than the discrete Fourier transform of correlator beamforming,
\begin{align}
  \widetilde{\mathcal{I}}_m^{\alpha\beta}(\hat{\boldsymbol{s}}_k) &= \sum_j \delta^2 \Delta\boldsymbol{r}_j \, e^{i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\Delta\boldsymbol{r}_j} \nonumber\\
  &\quad \left(\sum_{a_m,b_m} \sum_p \widetilde{B}_{a_m b_m}^{\alpha\beta pq*}(\Delta\boldsymbol{r}_j-\Delta\boldsymbol{r}_{a_m b_m}) \, \widetilde{V}_{a_m b_m}^{pq} \right) \, . \label{eqn:intra-station-pol-img-xfft-expl}
\end{align}
The gridding and Fourier transform are represented by the parenthesis and outer summation, respectively.

Mathematically, this is the counterpart to the operations performed on calibrated electric fields from the elements in EPIC, with few key differences -- (1) the calibrated visibilities can be accumulated and allowed for Earth rotation and bandwidth synthesis, (2) the gridding and spatial Fourier transform operates on visibilities rather than electric fields, and (3) the spatial Fourier transform can occur on a slower cadence of $t_\textrm{acc}$ or the calibration timescale, whichever is shorter. This is an explicit form of the optimal map-making (OMM) formalism \citep{Tegmark1997a} provides a lossless and optimal (least-squares) solution, which in the radio interferometric context is often referred to as the ``dirty image'' \citep{TMS2017,SIRA-II}. The purpose of $\widetilde{B}_{a_m b_m}^{\alpha\beta pq*}(\Delta\boldsymbol{r}_j-\Delta\boldsymbol{r}_{a_m b_m})$ is similar \citep{Morales+2009,Bhatnagar+2008} to its counterpart gridding operator in the EPIC architecture \citep{Morales2011}.

@@

Figure~\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I} shows the computational cost density breakdown for correlator FFT at the station level. The correlation and temporal averaging in Equation~(\ref{eqn:intra-station-pol-visibilities}) are the same as the correlator beamformer, scaling as $N_\textrm{a/s}(N_\textrm{a/s}-1)/2$ and has no dependence on $D_\textrm{a}$ or $D_\textrm{s}$. Hence, their per voxel costs scale as $(D_\textrm{s}/D_\textrm{a})^{-2}$. The gridding in Equation~(\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}) also scales as $N_\textrm{a/s}(N_\textrm{a/s}-1)/2$, but being independent of grid dimensions, its cost per voxel scales as $(D_\textrm{s}/D_\textrm{a})^{-2}$. The two-dimensional FFT in Equation~(\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}) depends only on the grid size and is independent of $N_\textrm{a/s}$. The FFT scales as $(D_\textrm{s}/D_\textrm{a})^2\log_2(D_\textrm{s}/D_\textrm{a})^2$ and thus its cost per voxel scales only weakly as $\log_2(D_\textrm{s}/D_\textrm{a})$. The correlation and temporal averaging have to be performed at every interval of $\delta t$ and are therefore, independent of $t_\textrm{acc}$, whereas the gridding and FFT need to be performed at a cadence of $t_\textrm{acc}$ and thus scale as $t_\textrm{acc}^{-1}$. For the chosen parameters of \texttt{LAMBDA-I} and $t_\textrm{acc}=1$~ms, the computational cost density is dominated by the cost of the gridding, followed by the correlator, temporal averaging of correlations, and FFT. However, for imaging on a slower cadence, $t_\textrm{acc}\gtrsim 5-10$~ms, the cost of correlation begins to dominate.

@@

The variation in the cost per voxel for $t_\textrm{acc}=1$~ms is dominated by the gridding cost, displaying the same trends observed in Figure~\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}. Figure~\ref{fig:multidim-incoherent-relative-compcost-XFFT-LAMBDA-I} shows the ratio of the computational cost of the XFFT architecture relative to the voltage beamforming architecture in Figure~\ref{fig:multidim-incoherent-compcost-BF-LAMBDA-I}. It is clearly noted that the cost of the XFFT architecture is less expensive than the BF in most of the parameter space, particularly for larger $D_\textrm{s}/D_\textrm{a}$. For the chosen \texttt{LAMBDA-I} parameters (gray dashed lines), the XFFT cost is roughly equal to that of voltage beamforming.

@@

The pointing direction of the virtual synthesised telescope from the elements of station, $m$ is denoted by $\hat{\boldsymbol{s}}_k$. The field of view and angular resolution of the intra-station $\widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k)$ are determined by the station element and the station dimensions, respectively. Using these electric fields as measurements from virtual synthesised telescopes, the inter-station imaging architecture can take any of the architectures that were already discussed under intra-station imaging by combining the electric fields from synthesised telescopes from different stations virtually pointed in the same direction. Since the computational costs depend on the station and array layouts, this necessitates hybrid architectures appropriate for multi-scale aperture arrays.

@@

For intra-station processing, EPIC emerges as the most viable among various architectures for all arrays considered on all timescales, except for \texttt{CASPA} when $t_\textrm{acc}\gtrsim 100$~ms where the XFFT architecture emerges to be more efficient than EPIC. This is because the stations having a sufficiently large number of elements packed densely is a predominant aspect of all these arrays (see Table~\ref{tab:array_params}. For arrays like \texttt{CASPA} that have lesser number of elements and/or are not as densely packed, the cadence interval becomes an important factor. For cadence intervals, $t_\textrm{acc}\gtrsim 100$~ms, the XFFT architecture becomes advantageous.

@@

Explosive transient phenomena and observational cosmology are two high impact science drivers in radio astronomy. The paradigm of aperture arrays used in radio astronomy is undergoing a transformation towards large numbers of low-cost elements grouped together in a hierarchy of spatial scales to achieve the desired science outcomes. Such a paradigm comes along with enormous challenges in computing and throughput because of the large numbers of elements spanning a multiplicity of spatial scales, and range of cadence intervals, real-time processing, and continuous operation. Different processing architectures are known to be optimal for specific layouts and processing cadences. However, a single architectural may not be optimal for a hierarchical spatial layout.

>>>>

\begin{abstract}
Two major areas of modern radio astronomy, namely, explosive astrophysical transient phenomena and observational cosmology, are driving aperture arrays towards large-$N$, low-cost element design consisting of multiple spatial scales. Multi-scale, hierarchical aperture arrays require a combination of data processing architectures-- pre-correlation beamformer, generic version of FFT-based direct imager, post-correlation beamformer, and post-correlation FFT imager-- operating on different spatial scales to obtain optimal performance in imaging the entire field of view. We use the metric of computational cost density defined over the dimensions of discovery space, namely, field of view, angular resolution, polarisation, frequency, and time to determine the most efficient hybrid architecture for aperture arrays spanning multiple spatial scales like the element size, stations (a collection of elements), and the array (a collection of stations) over a range of number of elements per station and number of stations. We use nominal parameters of specific upcoming and planned arrays-- Square Kilometre Array (SKA) at low frequencies (\texttt{SKA-low}), \texttt{SKA-low-core}, a long baseline extension to SKA-low (\texttt{LAMBDA-I}), compact all-sky phased array (\texttt{CASPA}), and a lunar array (\texttt{FarView-core})-- to determine the most optimal architectural hierarchy for each from a computational standpoint, and provide a guide for designing hybrid architectures for multi-scale aperture arrays. For large-$N$, dense-packed layouts, a FFT-based direct imager is most efficient for most cadence intervals, and for other layouts that have relatively lesser number of elements or greater sparsity in distribution, the best architecture is more sensitive to the cadence interval.
\end{abstract}

@@

Here we describe the first stage, namely, intra-station architectures that form the basis for both approaches. This first stage essentially synthesises station-scale apertures with simultaneous, multiple virtual pointings. These station-scale synthesised apertures from the first stage will act as building blocks for the second stage of inter-station processing. We will use $a_m$ to denote element, $a$, in station, $m$. A summary of these costs are provided in the ``intra-station'' column of Table~\ref{tab:intra-inter-station-coherent-imaging}, with a detailed discussion here.

@@

Figure~\ref{fig:1D-incoherent-compcost-BF-LAMBDA-I} shows the breakdown of various components of the computational cost per station per voxel for voltage beamforming as a function of the different station parameters. In each panel, all parameters except the one on the corresponding $x$-axis are kept fixed at the characteristic values of the \texttt{LAMBDA-I} station as highlighted in cyan. The total cost is dominated by the beamforming in Equation~(\ref{eqn:intra-station-pol-hol-img-expl}) denoted by black dashed lines. Beamforming cost per pixel scales linearly with the number of elements per station. The costs of squaring and temporal averaging in Equation~(\ref{eqn:intra-station-opt-pol-img-outprod}) are subdominant. Because they are performed on a per-pixel basis and not dependent on $N_\textrm{a/s}$, they remain constant across $N_\textrm{a/s}$. And because the number of independent pixels covering the field of view scales as $n_\textrm{bs} \simeq (D_\textrm{s}/D_\textrm{a})^2$, the cost per pixel remains constant across $D_\textrm{s}$ and $D_\textrm{a}$. All these operations have to be performed at every time step, $\delta t$, and hence remain constant regardless of $t_\textrm{acc}$. Although spatial averaging across multiple stations (orange dot-dashed line) is not a station-level operation, it is shown only to emphasise that it constitutes only a negligible cost relative to the other components. Note that it needs to be only performed on a slower cadence of $t_\textrm{acc}$ on a per-pixel basis, and thus scales inversely with $t_\textrm{acc}$.

@@

Figure~\ref{fig:multidim-incoherent-compcost-BF-LAMBDA-I} shows a covariant view of the total computational cost density per station per voxel for the beamforming architecture taken two station parameters at a time while keeping the rest fixed at the nominal values of a \texttt{LAMBDA-I} station denoted by grey dashed lines. The variation in the cost per voxel is dominated by the beamforming cost and exhibits the same trends observed in Figure~\ref{fig:1D-incoherent-compcost-BF-LAMBDA-I}, namely, linear in $N_\textrm{a/s}$ while remaining constant with $D_\textrm{a}$ and $D_\textrm{s}$.

@@

Forming independent beams that simultaneously fill the entire field of view can be achieved by a two-dimensional spatial Fast Fourier Transform (FFT) of the measurements on the aperture plane. This can be enabled even for an irregular layout of station elements by gridding the calibrated electric fields measured by the station elements. The Optimal Map Making formalism \citep[OMM;][]{Tegmark1997a} formalism adopted by \citet{Morales2011} is the basis of the architecture called E-field Parallel Imaging Correlator \cite[EPIC;][]{Thyagarajan+2017},
\begin{align}
    \widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k) &= \sum_j \delta^2 \boldsymbol{r}_j \, e^{i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\boldsymbol{r}_j} \left(\sum_{a_m} \sum_p \widetilde{W}_{a_m}^{\alpha p*}(\boldsymbol{r}_j-\boldsymbol{r}_{a_m}) \, \widetilde{E}_{a_m}^p \right) \, . \label{eqn:intra-station-pol-hol-img-epic}
\end{align}
The expression within the parenthesis represents the gridding operation, and the outer summation denotes the Fourier transform implemented via FFT.

@@

Application of the FFT will have the effect of simultaneously beamforming over the entire field of view, $\Omega_\textrm{a}$. The convolution with $\widetilde{W}_{a_m}^{p\alpha*}(\boldsymbol{r})$ has two purposes. Firstly, it can be chosen to optimise specific properties in the synthesised image. For example, choosing $\widetilde{W}_{a_m}^{p\alpha}(\boldsymbol{r})=W_{a_m}^{p\alpha*}(\boldsymbol{r})$ will optimise the signal-to-noise ratio in the image. Secondly, it also acts as gridding convolution transforming data from discrete and arbitrary locations onto a regular grid, thereby facilitating the application of a two-dimensional spatial FFT. Finally, the polarised intensities are estimated according to Equation~(\ref{eqn:intra-station-opt-pol-img-outprod}).

@@

Figure~\ref{fig:1D-incoherent-compcost-EPIC-LAMBDA-I} shows a breakdown of the costs per station per voxel of various components constituting the EPIC architecture, with nominal values of a \texttt{LAMBDA-I} station marked in cyan. The dominant cost is from the two-dimensional FFT (black dashed lines) in Equation~\ref{eqn:intra-station-pol-hol-img-epic}.
% or equivalently, Equation~\ref{eqn:intra-station-pol-hol-img-epic-LA}.
Because the spatial FFT cost depends only on the grid dimensions, it is independent of $N_\textrm{a/s}$ as the grid can hold any $N_\textrm{a/s}$ elements as physically possible. This spatial FFT scales roughly as $(\gamma_\textrm{s} \, D_\textrm{s}/D_\textrm{a})^2\log_2(\gamma_\textrm{s}\, D_\textrm{s}/D_\textrm{a})^2$, where, $\gamma_\textrm{s}$ is a constant padding factor. Notably, because the spatial FFT produces an image over the entire field of view at every independent pixel instantly, it is not a pixelwise operation, and therefore, its cost per pixel is $\sim \log_2(D_\textrm{s}/D_\textrm{a})$, showing a weak dependence on $D_\textrm{s}$ and $D_\textrm{a}$. The pixelwise squaring (pink dot-dashed lines) and temporal averaging (blue dot-dashed lines) costs in Equation~\ref{eqn:intra-station-opt-pol-img-outprod}
% {eqn:intra-station-opt-polarimetric-direct-img-LA}
are subdominant. These being per pixel operations, the cost per voxel remains constant with $N_\textrm{a/s}$, $D_\textrm{s}$ and $D_\textrm{a}$ similar to the voltage beamforming architecture. The gridding (black dotted lines) with weights in Equation~(\ref{eqn:intra-station-pol-hol-img-epic})
% and (\ref{eqn:intra-station-pol-hol-img-epic-LA})
is a per-element operation and depends only on $N_\textrm{a/s}$, it scales linearly with $N_\textrm{a/s}$ and inverse squared with $D_\textrm{s}$ per voxel. Overall, the gridding cost is significantly subdominant. Again, the spatial averaging of pixels across multiple stations (orange dot-dashed line) is not a station-level operation, but is only shown to emphasise that it operates on a much slower cadence of $t_\textrm{acc}$, scaling inversely with $t_\textrm{acc}$, and adds only negligible cost to the overall cost budget.

@@

\begin{figure*}
\centering
\subfloat[][Covariant cost density of EPIC architecture \label{fig:multidim-incoherent-compcost-EPIC-LAMBDA-I}]{\includegraphics[width=0.48\textwidth]
% {figures/multidim_epic_per_pixel_cost_analysis_LAMBDA-I.pdf}
{fig_05a.pdf}  
}
\subfloat[][Covariant cost of EPIC architecture relative to BF architecture \label{fig:multidim-incoherent-relative-compcost-EPIC-LAMBDA-I}]{\includegraphics[width=0.48\textwidth]
% {figures/multidim_epic_relative_cost_analysis_LAMBDA-I.pdf}
{fig_05b.pdf}
} 
\caption{(Left): Two-dimensional slices of the computational cost per station per voxel for coherent-incoherent imaging using station-level EPIC. Because the spatial FFT dominates the overall cost budget, the computational cost density is relatively insensitive to $N_\textrm{a/s}$, $D_\textrm{s}$, and $D_\textrm{a}$. (Right): Ratio of computational cost of EPIC to voltage beamforming. For $N_\textrm{a/s}\gtrsim 100$, EPIC has a significant advantage over voltage beamforming. Cyan contours indicate nominal values for a \texttt{LAMBDA-I} station. \label{fig:incoherent-compcost-EPIC-LAMBDA-I}}
\end{figure*}

@@

Figure~\ref{fig:multidim-incoherent-compcost-XBF-LAMBDA-I} is a covariant view of the total computational cost density per station per voxel for the correlator beamformer architecture taken two station parameters at a time while keeping the rest fixed at the nominal values of a \texttt{LAMBDA-I} station denoted by grey dashed lines. The variation in the cost per voxel for $t_\textrm{acc}=1$~ms is dominated by the spatial DFT cost which is relatively constant in $D_\textrm{a}$ and $D_\textrm{a}$, and scales as $N_\textrm{a/s}^2$, displaying the same trends observed in Figure~\ref{fig:1D-incoherent-compcost-XBF-LAMBDA-I}. Figure~\ref{fig:multidim-incoherent-relative-compcost-XBF-LAMBDA-I} shows the ratio of the computational cost of the XBF architecture relative to the voltage beamforming architecture in Figure~\ref{fig:multidim-incoherent-compcost-BF-LAMBDA-I}. It is clear that the cost of the XBF architecture is less expensive only for $N_\textrm{a/s}\lesssim 100$, and gets worse with increasing $N_\textrm{a/s}$.

@@

In this approach, $\widetilde{V}_{a_m b_m}^{pq}$, the calibrated and inverse noise covariance weighted station visibilities in station $m$ in Equation~(\ref{eqn:intra-station-pol-visibilities}) are gridded with weights, $\widetilde{B}_{a_m b_m}^{\alpha\beta;pq*}(\Delta\boldsymbol{r}_j-\Delta\boldsymbol{r}_{a_m b_m})$, onto a grid and Fourier transformed using an FFT rather than the discrete Fourier transform of correlator beamforming,
\begin{align}
  \widetilde{\mathcal{I}}_m^{\alpha\beta}(\hat{\boldsymbol{s}}_k) &= \sum_j \delta^2 \Delta\boldsymbol{r}_j \, e^{i\frac{2\pi}{\lambda} \hat{\boldsymbol{s}}_k\cdot\Delta\boldsymbol{r}_j} \nonumber\\
  &\quad \left(\sum_{a_m,b_m} \sum_p \widetilde{B}_{a_m b_m}^{\alpha\beta;pq*}(\Delta\boldsymbol{r}_j-\Delta\boldsymbol{r}_{a_m b_m}) \, \widetilde{V}_{a_m b_m}^{pq} \right) \, . \label{eqn:intra-station-pol-img-xfft-expl}
\end{align}
The gridding and Fourier transform (using FFT) are represented by the parenthesis and outer summation, respectively.

Mathematically, this is the counterpart to the operations performed on calibrated electric fields from the elements in EPIC, with few key differences -- (1) the calibrated visibilities can be accumulated and allowed for Earth rotation and bandwidth synthesis, (2) the gridding and spatial Fourier transform operate on visibilities rather than electric fields, and (3) the spatial Fourier transform can occur on a slower cadence of $t_\textrm{acc}$ or the calibration timescale, whichever is shorter. This is an explicit form of the optimal map-making (OMM) formalism \citep{Tegmark1997a}, and provides a lossless and optimal (least-squares) solution, which in the radio interferometric context is often referred to as the ``dirty image'' \citep{TMS2017,SIRA-II}. The purpose of $\widetilde{B}_{a_m b_m}^{\alpha\beta;pq*}(\Delta\boldsymbol{r}_j-\Delta\boldsymbol{r}_{a_m b_m})$ is similar \citep{Morales+2009,Bhatnagar+2008} to its counterpart gridding operator in the EPIC architecture \citep{Morales2011}.

@@

Figure~\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I} shows the computational cost density breakdown for XFFT at the station level. The correlation and temporal averaging in Equation~(\ref{eqn:intra-station-pol-visibilities}) are the same as the correlator beamformer, scaling as $N_\textrm{a/s}(N_\textrm{a/s}-1)/2$ and has no dependence on $D_\textrm{a}$ or $D_\textrm{s}$. Hence, their per voxel costs scale as $(D_\textrm{s}/D_\textrm{a})^{-2}$. The gridding in Equation~(\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}) also scales as $N_\textrm{a/s}(N_\textrm{a/s}-1)/2$, but being independent of grid dimensions, its cost per voxel scales as $(D_\textrm{s}/D_\textrm{a})^{-2}$. The two-dimensional FFT in Equation~(\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}) depends only on the grid size and is independent of $N_\textrm{a/s}$. The FFT scales as $(D_\textrm{s}/D_\textrm{a})^2\log_2(D_\textrm{s}/D_\textrm{a})^2$, and thus its cost per voxel scales only weakly as $\log_2(D_\textrm{s}/D_\textrm{a})$. The correlation and temporal averaging have to be performed at every interval of $\delta t$ and are therefore, independent of $t_\textrm{acc}$, whereas the gridding and FFT need to be performed at a cadence of $t_\textrm{acc}$, and thus scale as $t_\textrm{acc}^{-1}$. For the chosen parameters of \texttt{LAMBDA-I} and $t_\textrm{acc}=1$~ms, the computational cost density is dominated by the cost of the gridding, followed by the correlator, temporal averaging of correlations, and FFT. However, for imaging on a slower cadence, $t_\textrm{acc}\gtrsim 5-10$~ms, the cost of correlation begins to dominate.

@@

The variation in the cost per voxel for $t_\textrm{acc}=1$~ms is dominated by the gridding cost, displaying the same trends observed in Figure~\ref{fig:1D-incoherent-compcost-XFFT-LAMBDA-I}. Figure~\ref{fig:multidim-incoherent-relative-compcost-XFFT-LAMBDA-I} shows the ratio of the computational cost of the XFFT architecture relative to the voltage beamforming architecture in Figure~\ref{fig:multidim-incoherent-compcost-BF-LAMBDA-I}. Clearly, the cost of the XFFT architecture is less expensive than the BF in most of the parameter space, particularly for larger $D_\textrm{s}/D_\textrm{a}$. For the chosen \texttt{LAMBDA-I} parameters (gray dashed lines), the XFFT cost is roughly equal to that of voltage beamforming.

@@

The pointing direction of the virtual synthesised telescope from the elements of station, $m$ is denoted by $\hat{\boldsymbol{s}}_k$. The field of view and angular resolution of the intra-station $\widetilde{\mathcal{E}}_m^\alpha(\hat{\boldsymbol{s}}_k)$ are determined by the station element and the station dimensions, respectively. Using these electric fields as measurements from virtual synthesised telescopes, the inter-station imaging architecture can take any of the architectures that were already discussed under intra-station imaging by combining the electric fields from synthesised telescopes from different stations virtually pointed in the same direction. Since the computational costs depend on the station and array layouts, this necessitates hybrid architectures appropriate for multi-scale aperture arrays. A summary of these costs are provided in the ``inter-station'' column of Table~\ref{tab:intra-inter-station-coherent-imaging}, with a detailed discussion below.

@@

For intra-station processing, EPIC emerges as the most viable among various architectures for all arrays considered on all timescales, except for \texttt{CASPA} when $t_\textrm{acc}\gtrsim 100$~ms where the XFFT architecture emerges to be more efficient than EPIC. This is because the stations having a sufficiently large number of elements packed densely is a predominant aspect of all these arrays (see Table~\ref{tab:array_params}). For arrays like \texttt{CASPA} that have lesser number of elements and/or are not as densely packed, the cadence interval becomes an important factor. For cadence intervals, $t_\textrm{acc}\gtrsim 100$~ms, the XFFT architecture becomes advantageous.

@@

Explosive transient phenomena and observational cosmology are two high impact science drivers in radio astronomy. The paradigm of aperture arrays used in radio astronomy is undergoing a transformation towards large numbers of low-cost elements grouped together in a hierarchy of spatial scales to achieve the desired science outcomes. Such a paradigm comes along with enormous challenges in computing and throughput because of the large numbers of elements spanning a multiplicity of spatial scales, a range of cadence intervals, real-time processing, and continuous operation model. Different processing architectures are known to be optimal for specific layouts and processing cadences. However, a single architectural may not be optimal for a hierarchical spatial layout.

This paper explores hybrid architecture solutions for existing and planned multi-scale aperture arrays (\texttt{SKA1-low-core}, \texttt{SKA1-low}, \texttt{LAMBDA-I}, \texttt{CASPA}, and \texttt{FarView-core}) on a wide range of cadence intervals, 0.1~ms$\le t_\textrm{acc} \le 10$~s, over the full field of view, using the metric of computational cost density evaluated over the discovery parameter space. The interferometric data processing architectures considered here are pre-correlation voltage beamformer (BF), a generic FFT-based direct imager (EPIC), beamforming from correlations (XBF), and FFT-based imaging of correlations (XFFT). The computational cost budget for various architecture combinations for these different arrays are broken down into their detailed components.

For densely packed layouts with large numbers of elements, EPIC is computationally most efficient regardless of processing cadence. This was the case for the station layouts of all the arrays considered, and hence EPIC emerges as the computationally most efficient architecture for full field of view, station level data processing on all cadence intervals except for \texttt{CASPA} on $t_\textrm{acc}\gtrsim 100$~ms, when XFFT becomes more efficient. 
% Thus, in almost cases of station layouts considered here, EPIC appears to hold a clear advantage for full field of view, intra-station processing.
